---
title: "Publications and other Scientific Communications"
description: |
  A list of my publications and presentations (talks and posters), as well as content for scientific communications that include videos, community outreach, and more.
output: 
  distill::distill_article:
    toc: true
    toc_float: true #doesn't really work...
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


##### Libraries
library("vembedr")


##### Number counters

#Counter for publications
mycounter_pubs <- 0

#Counter for presentation talks
mycounter_talks <- 0

#Counter for presentation posters
mycounter_posters <- 0

#Counter for videos
mycounter_videos <- 0



#Adding `&emsp;` is four spaces
#Addint `&ensp;` is two spaces
```


# Publications

`r mycounter_pubs <- mycounter_pubs +1; mycounter_pubs`. &emsp; Hegarty, M., He, C., Boone, A.P., **Yu, S.**, Jacobs, E.G., & Chrastil, E.R. (2022). Understanding Differences in Wayfinding Strategies. *Topics in Cognitive Science*. https://doi.org/10.1111/tops.12592.

`r mycounter_pubs <- mycounter_pubs +1; mycounter_pubs`. &emsp; **Yu, S.**, Boone, A.P., He, C., Davis, R.C., Hegarty, M., Chrastil, E.R., & Jacobs, E.G. (2021). Age-Related Changes in Spatial Navigation Are Evident by Midlife and Differ by Sex. *Psychological Science*, 32(5), 692-704. https://doi.org/10.1177/0956797620979185. **_Featured in Association for Psychological Science's "This Week in Psychological Science" exclusive newsletter._**

`r mycounter_pubs <- mycounter_pubs +1; mycounter_pubs`. &emsp; Pritschet, L., Santander, T., Taylor, C.M., Layher, E., **Yu, S.**, Miller, M.B., Grafton, S.T., & Jacobs, E.G. (2020). Functional reorganization of brain networks across the human menstrual cycle. *NeuroImage*, 220, 117091. https://doi.org/10.1016/j.neuroimage.2020.117091.

`r mycounter_pubs <- mycounter_pubs +1; mycounter_pubs`. &emsp; Taylor, C.M., Pritschet, L., **Yu, S.**, & Jacobs, E.G. (2019). Applying a Women’s Health Lens to the Study of the Aging Brain. *Frontiers in human neuroscience*, 13, 224. https://doi.org/10.3389/fnhum.2019.00224.

`r mycounter_pubs <- mycounter_pubs +1; mycounter_pubs`. &emsp; Suarez, N.A., Samano, M., **Yu, S.**, Snyder, S. & Chukoskie, L. (2019). Research-practice partnerships in STEM informal learning environments. In P.K. Kuhl, S-S. Lim, S. Guerriero & D. Van Damme (Eds.), *Developing Minds in the Digital Age: Towards a Science of Learning for 21st Century Education*, (pp. 187-195). Paris, France: Educational Research and Innovation, OECD Publishing. https://doi.org/10.1787/562a8659-en.


# Presentations

## Talks

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.** (September 2021). Sex Differences and Aging in Spatial Navigation. *Lunch & Learn Series*. UCSB Graduate Division, Santa Barbara, CA, USA. **_Highlighted as "the most accessible STEM presentation seen at Lunch & Learn" by the Assistant Director of Professional Development for Graduate Division._** _See <a href="#videos">Videos</a> to watch talk._

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.**, Boone, A.P., He, C., Davis, R., Hegarty, M., Chrastil, E.R., & Jacobs, E.G. (October 2020). Sex Differences and Age-Related Differences in Spatial Navigation. *3rd Interdisciplinary Navigation Symposium (iNAV)*. Virtual Presentation, Cortina d'Ampezzo, Italy.

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.** (May 2020). Spatial Navigation, Sex Differences, and Aging. *43rd Annual Minicon*. UCSB Department of Psychological and Brain Sciences, Santa Barbara, CA, USA.

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.** (March 2020). Why Spatial Navigation is Important. *Invited Guest Lecture for PSY 137: Behavioral Endocrinology*. UCSB Department of Psychological and Brain Sciences, Santa Barbara, CA, USA.

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.**, Pritschet, L., Taylor, C.M. & Jacobs, E.G. (February 2018). *Cognitive Neuroscience Seminar Series*. UCSB Department of Psychological & Brain Sciences, Santa Barbara, CA, USA.

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.**, Chu, D. & Nitz, D.A. (June 2016). Role of the Hippocampus and Parietal Cortex in Rule-Based Environmental Fragmentation. *Cognitive Science Undergraduate Honors Thesis Presentations*, La Jolla, CA, USA.

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.**, Smith, C.N. & Squire, L.R. (April 2016). Eye Movements, Memory, and Awareness of Whether a Scene Is Old or New. *29th Annual Undergraduate Research Conference*, La Jolla, CA, USA.

`r mycounter_talks <- mycounter_talks +1; mycounter_talks`. &emsp; **Yu, S.**, Chukoskie, L. & Townsend, J. (April 2015). Whatcha Looking At? Social and Non-Social Eye Fixations of Children with ASD in a Change Blindness Task. *28th Annual Undergraduate Research Conference*, La Jolla, CA, USA.



## Posters

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; **Yu, S.**, Boone, A.P., He, C., Davis, R.C., Hegarty, M., Chrastil, E.R., & Jacobs, E.G. (May 2020). Sex Differences and Age-Related Changes in Spatial Navigation. *2020 Organization for the Study of Sex Differences Annual Meeting*, Marina del Rey, CA, USA. _Abstract accepted but meeting canceled due to COVID-19._


`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; Layher, E., Pritschet, L., Santander, T., Taylor, C.M., **Yu, S.**, Grafton, S.T., Jacobs, E.G., & Miller, M.B. (June 2019). Within-subject fMRI analyses of the successful retrieval effect in recognition memory. *2019 Organization for Human Brain Mapping Annual Meeting*, Rome, Italy. 

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; Shelley, L.E., Chu, D., **Yu, S.**, Kim, S., Ragland, S., Miyamoto, K., Park, J. & Nitz, D.A. (October 2016). Creating spaces: A possible function of the hippocampus and parietal networks. *2nd Annual Frontiers of Innovation Scholars Program Symposium*, La Jolla, CA, USA. 

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; **Yu, S.**, Smith, C.N. & Squire, L.R. (May 2016). Eye Movements, Memory, and Awareness of Whether a Scene Is Old or New. *2016 Psychology Honors Poster Session*, La Jolla, CA, USA.

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; **Yu, S.**, Smith, C.N. & Squire, L.R. (February 2016). Differential eye movements for old and new scenes under free viewing conditions are unrelated to awareness and are hippocampus-independent. *2016 Inter-Science of Learning Center Conference*, Arlington, VA, USA. 

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; **Yu, S.**, Smith, C.N. & Squire, L.R. (January 2016). Differential eye movements for old and new scenes under free viewing conditions are unrelated to awareness and are hippocampus-independent. *Temporal Dynamics of Learning Center All Hands Meeting 2016*, La Jolla, CA, USA. 

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; Samano, M., Suarez, N.A., **Yu, S.** & Chukoskie, L. (November 2015). A Good Balance Between Research and Outreach in an Informal Science Learning Environment. *1st Annual Frontiers of Innovation Scholars Program Symposium*, La Jolla, CA, USA. 

`r mycounter_posters <- mycounter_posters +1; mycounter_posters`. &emsp; **Yu, S.**, Chukoskie, L. & Townsend, J. (May 2015). Whatcha Looking At? Social and Non-Social Eye Fixations of Children with ASD in a Change Blindness Task. *2015 Inter-Science of Learning Center Conference*, La Jolla, CA, USA. 


# Videos


`r mycounter_videos <- mycounter_videos +1; mycounter_videos`. &emsp; **Crossroads**

As part of my Crossroads fellowship, I worked with three graduate students and four professors to design an interdisciplinary pedagogy on the topics surrounding my academic research that investigates sex hormones, aging, and spatial navigation. The results were multi-part videos that explained why there are sex differences in spatial navigation from the perspective of four main disciplines: neuroendocrinology, cognitive psychology, human geography, and biological anthropology. The goal of the videos were to present them to undergraduate students and diverse audiences who would benefit from applying course concepts to tangible research, in addition to thinking critically about how science is contextualized across multiple disciplines.

I conceptualized the ideas of the teaching videos, designed and reviewed the layout of the presentations, directed the content and goals of the videos, wrote scripts, and did voice-overs for two of the videos. I was responsible for creating the video that explains the research question from the perspective of a neuroendocrinologist:


```{r echo=FALSE}
#Endocrinology video
embed_url("https://www.youtube.com/watch?v=Uo1YevDB6Q0")
```


To watch the other videos, please see the [playlist on YouTube](https://youtube.com/playlist?list=PLRk7eVChv4aQj-mFQKxl_OB5o5LDgiCl9). The recommended order to watch the videos are:

- Introduction

- Cognitive Psychology

- Endocrinology

- Biological Anthropology

- Human Geography


<br>

`r mycounter_videos <- mycounter_videos +1; mycounter_videos`. &emsp; **Lunch & Learn**

My recorded talk for UCSB Graduate Division's *Lunch & Learn Series* from September 2021.

Password: LunchLearn2021


```{r echo=FALSE}
#UCSB Grad Div Lunch and Learn
#Password: LunchLearn2021
embed_url("https://vimeo.com/620065554")
```


