---
title: "Spatial Analysis: California Oil Spills"
description: |
  In this project, I plotted spatial maps to visualize California oil spills by county and conducted point pattern analysis to investiate locations of spatial clusters of oil spills.
author:
  - name: Shuying Yu
citation_url: https://esswhy.github.io/projects/spatial_oil_spill/
date: 03-02-2022
categories:
  - Spatial Analysis
  - R
output:
  distill::distill_article:
    self_contained: false
    code_folding: Code
    toc: true
    toc_float: true
---

<style>
.html-widget {
    margin: auto;
}
</style>

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")

#Attach packages

#For spatial maps and stats
library(spatstat)
library(maptools)

#For general wrangling
library(tidyverse)
library(here)
library(janitor)

#Additional spatial maps
library(sf)
library(tmap)

#For plots
library(grid)
library(ggpubr)
library(gridExtra)
library(patchwork)
```


# Introduction

In this project, I used spatial data visualization and point pattern analysis to explore California oil spill incidents that were tracked and quantified in the Office of Spill Prevention and Response Incident Tracking Database System. Exploratory data analysis was performed and maps of oil spills were plotted in California by county. Point pattern analysis using *G*-function was conducted to test for complete spatial randomness.


# Research Question

Do oil spills in California tend to be more clustered or more uniformly distributed than complete spatial randomness? Using spatial visualizations and point pattern analysis, we aim to investigate whether oil spill incidents in California are either sporadic or clustered in specific spatial locations.


# Data and Methods

The California Department of Fish and Wildlife (CDFW) describes oil spill "incidents" as "a discharge or threatened discharge of petroleum or other deleterious material into the waters of the state." (CDFW, 2020). There were over 3,200 tracked oil spill incidents recorded in California during the year 2008. Data is made publicly available by the Office of Spill Prevention and Response (OSPR) Incident Tracking Database System (2009). The database contains statewide oil spill tracking information system with data collected by OSPR Field Response Team members for Marine oil spills and by OSPR Inland Pollution Coordinators and Wardens for Inland incidents (OSPR, 2009). Purposes of the data include program planning, public education, spill preparedness, and response performance.

Exploratory spatial analysis included plotting the locations of oil spill incidents by creating an interactive thematic map and assessing counts of oil spills by county using a choropleth map with shapefile courtesy of the California Open Data Portal (2016). Kernel desnsity plot was also presented. Spatial point pattern analysis was performed to assess spatial clustering of the locations of oil spill incidents. *G*-function was calculated as the proportion of oil spill incidents with the nearest neighbor within spatial unit distance. All analyses were conducted in R version 4.1.1 and RStudio version 1.4.1717.


# Results

## Locations of Oil Spills in California

### Basic Map

Using `geom_sf`, we can plot the oil spill incident locations using shapefiles, which are a common way to store geospatial data.

```{r}

########## Shapefile with CA county boundaries ##########

#Read in the county boundary shapefile
ca_counties_sf <- read_sf(here("_projects", "data", "ca_county_boundaries",
                               "CA_Counties_TIGER2016.shp")) %>% 
  
  #Clean variable names
  clean_names() %>% 
  
  #Select name and rename to county_name
  #Now sf file contains these two variables: this and geometry
  dplyr::select(county_name = name)


#Check the project for coordinate reference system (CRS)
#st_crs(ca_counties_sf) #WGS 84, EPSG 3857



########## Shapefile with oil spill data ##########

#Read in the oil spill data
ca_spills_sf <- read_sf(here("_projects", "data", "ds394", "ds394.shp")) %>% 
  
  #Clean variable names
  clean_names() %>% 
  
  #Rename variables
  rename(county_name = "localecoun",
         inland_marine = "inlandmari") 


#Now convert using EPSG code
ca_spills_sf2 <- st_transform(ca_spills_sf, 3857)


#Check the project for coordinate reference system (CRS)
#st_crs(ca_spills_sf2) #WGS 84, EPSG 3857



########## Plot location of oil spill incidents

#Visualize sf objects
ggplot() +
  
  #Plot CA county map
  geom_sf(data = ca_counties_sf,
          fill = "lightblue") +
  
  #Plot CA oil spills
  #Change size, color, transparency of points
  geom_sf(data = ca_spills_sf2, size = 1,
          color = "darkorange", alpha = 0.6) +
  
  #Add x-axis and y-axis labels
  labs(x = "\nLatitude",
       y = "Longitude\n") +
  
  #Change theme
  theme_minimal() +
  
  #Add custom theme
  #Change size of axis tick text
  theme(axis.text = element_text(size = 10),
        
        #Bold and change size of axes labels
        axis.title = element_text(face = "bold", size = 12))
```
**Figure 1.** Map of California oil spill incidents (orange dots) that were tracked and quantified in the year 2008. Data source: OSPR (2009).

<br>

### Interactive Thematic Map

Using `tmap`, we can plot the oil spill incident locations in an interactive thematic map. The locations of the oil spills will show whether the event took place in marine or inland environments.

```{r}
#Rename files for county boundaries
County <- ca_counties_sf

#Rename files for oil spills
`Oil Spills` <- st_transform(ca_spills_sf, 3857) %>% 
  
  #Rename variable name for sake of showing as 
  rename(Environment = "inland_marine")



#View interactive tmap
tmap_mode("view")

#Plot map
#Change layers to view
  tm_basemap(c(`Street Map` = "OpenStreetMap",
               `Topographic Map` = "OpenTopoMap")) +
    
  #Add county boundaries
  tm_shape(County) +

  #Change transparency of polygons of counties
  tm_polygons(alpha = 0) +
    
  #Add oil spill data
  tm_shape(`Oil Spills`) +
  
  #Plot oil spills based on inland vs. marine
  tm_dots("Environment")
```


## Frequency of Oil Spills by County

### Choropleth Map

```{r preview=TRUE}

########## Join datasets and fix NA data ##########

#Use `st_join` for spatial joins
ca_counties_oil_sf <- ca_counties_sf %>% 
  st_join(ca_spills_sf2)

#Check
#Modoc county is NA, so manually enter in 0
# test <- ca_counties_oil_sf %>% 
#   filter(is.na(inland_marine) == TRUE)
  


########## Find counts of inland oil spill events by county ##########

ca_oil_counts_sf <- ca_counties_oil_sf %>% 
  
  #Filter for inland spills and any NA data
  filter(inland_marine %in% c("Inland", NA)) %>% 
  
  #Rename county_name variable
  #Same as county_name.y
  rename("county_name" = county_name.x) %>% 
  
  #Group by county
  group_by(county_name) %>% 
  
  #Sum of records/oil spills
  summarize(n_records = sum(!is.na(county_name)))





########## Plot choropleth map of oil spills by county ##########

ggplot(data = ca_oil_counts_sf) +
  
  #Fill with number of oil spills
  #Change border color and size
  geom_sf(aes(fill = n_records), color = "white", size = 0.1) +
  
  #Change gradient colors
  #NA values will be lightgrey
  scale_fill_gradientn(colors = c("lightgrey", "orange", "red"),
                       na.value = "lightgrey") +
  
  #Change theme
  theme_minimal() +
  
  #Change x-axis and y-axis
  #Change legend title
  labs(x = "\nLatitude",
       y = "Longitude\n",
       fill = "Number of \nInland Oil Spills") +
  
  #Add custom theme
  #Change size of axis tick text
  theme(axis.text = element_text(size = 10),
        
        #Bold and change size of axes labels
        axis.title = element_text(face = "bold", size = 12))
```
**Figure 2.** Choropleth map of California inland oil spill incidents recorded in the year 2008. Gradient colors represent number of oil spills from lowest (grey) to highest (red). Data source: OSPR (2009).

<br>

Based on the choropleth map of California inland oil spill incidents (**Figure 2**), Los Angeles County has the highest number of recorded oil spill incidents in the year 2008. From the map of the location of oil spill incidents (**Figure 1**), it appears that most oil spills were clustered in Los Angeles County as well as some counties in the Bay Area.


## Point Pattern Analysis

### Kernel Density Map

To further determine whether oil spills tend to be more clustered or are spatially random, we will perform a point pattern analysis on the data. First, we can make a kernel density map of the location of oil spill incidents, focusing on Los Angeles County.


```{r}

########## Isolate LA County in shapefile map and oil spill data ##########

#Plot just oil spills of LA County
la_county_sf <- ca_counties_sf %>% 
  filter(county_name == "Los Angeles")

la_county_spills <- ca_spills_sf2 %>% 
  filter(county_name == "Los Angeles")



########## Plot oil spill in just LA County ##########

#Visualize sf objects
la_spill_plot <- ggplot() +
  
  #Plot LA county map
  geom_sf(data = la_county_sf,
          fill = "lightblue") +
  
  #Plot LA oil spills
  #Change size, color, transparency of points
  geom_sf(data = la_county_spills, size = 1,
          color = "darkorange", alpha = 0.6) +
  
  #Add x-axis and y-axis labels
  #Add tag
  labs(
       #Remove to combine x-axis together
       #x = "\nLatitude",
       y = "Longitude\n",
       tag = "A") +
  
  #Change theme
  theme_minimal() +
  
  #Change size of axis tick text
  theme(axis.text = element_text(size = 10),
        
        #Tilt x tick
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        
        #Bold and change size of axes labels
        axis.title = element_text(face = "bold", size = 12))



########## Convert files to spatial and spatial point pattern objects ##########

#Convert to object 'Spatial'
oil_sp  <- as(la_county_spills,"Spatial") 

#Convert to spatial point pattern
oil_ppp <- as(oil_sp, "ppp") 

#Convert to object 'Spatial1
la_sp  <- as(la_county_sf, "Spatial")

#Comes from spatial stats package: ppp and owin
#Convert to spatial point pattern from `spatstat` package
la_win <- as(la_sp, "owin") 
 
#Combine as a point pattern object (points + window):
oil_full <- ppp(oil_ppp$x, oil_ppp$y, 
                window = la_win)

#Check where points lie within county border
#plot(oil_full)



########## Kernel density plot ##########

#Take points and say x,y position spatial points have a probability distribution around it
#Sigma=1, units of reference sys is in meters, create normal curve where curve is and give std of 1 meter around that
oil_density <- density(oil_full, sigma = 15000)

#Convert density_spatstat into a stars object
density_stars <- stars::st_as_stars(oil_density)

#Convert density_stars into an sf object
density_sf <- st_as_sf(density_stars) %>%
  st_set_crs(3857)


#Plot density map
la_density_plot <- ggplot() +
  
  #Plot density object, fill with density `v`
  geom_sf(data = density_sf, aes(fill = v), col = NA) +
  
  #Choose gradient scale color
  #Missing data just colored as blue
  scale_fill_gradientn(colours = c("blue", "red", "yellow"), 
                       na.value = "blue") +
  
  #Plot LA county map
  geom_sf(data = st_boundary(la_county_sf)) +
  
  #Plot LA oil spills
  #Change size, color, transparency of points
  # geom_sf(data = la_county_spills, size = 1, col = "darkorange",
  #         alpha = 0.6) +
  
  #Add x-axis labels
  #Change legend title, add tag
  labs(
       #Remove to combine x-axis together
       #x = "\nLatitude",
       fill = "Density of\nOil Spills",
       tag = "B") +
  
  #Change theme
  theme_minimal() +
  
  #Change size of axis tick text
  theme(axis.text = element_text(size = 10),

        #Tilt x tick
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))



########## Combine plots together ##########

#Plot together using `patchwork`
result <- la_spill_plot + la_density_plot
point_kernel <- patchwork::patchworkGrob(result)

#Add x-axis title
#Change size, make bold, adjust vertical placement
xtitle <- ggpubr::text_grob("Latitude", size = 12, face = "bold",
                            vjust = -0.3)

#Plot figure and axis together using `gridExtra`
gridExtra::grid.arrange(point_kernel, 
                        bottom = xtitle) 
```

**Figure 3.** Map for the location (**A**) and the kernel density (**B**) of oil spill incidents (orange dots) for Los Angeles County in the year 2008. Density of oil spill incidents in Los Angeles County are represented from lowest (dark blue) to highest (yellow). Points outside of the county boundaries are oil spill incidents that have occurred in marine environments rather than inland environments. Data source: OSPR (2009).

<br>

### Nearest Neighbor using G-Function

We will use the *G*-function to compare our observed point pattern to a simulated complete spatial randomness scenario, to help us determine if it is more clustered or more uniform than complete spatial randomness. Units of *G(r)* are in degrees lat-long.

```{r}
#Make a sequence of distances over which to calculate G(r)
r_vec <- seq(0, 10000, by = 100) 

#Calculate the actual and theoretical G(r) values, using 100 simulations of CRS for the "theoretical" outcome

#Envelope function within which we run simulations for CSR, and calculate the G-function value at distances r for each simulation
gfunction <- envelope(oil_full, 
                      
                      #Use G-function
                      fun = Gest, 
                      
                      #Sequence of distances over which we’ll calculate 
                      #the proportion of points with nearest neighbor within that range
                      r = r_vec, 
                      
                      #100 simulations of CSR
                      nsim = 100, 
                      
                      #2nd highest and 2nd lowest values from simulations 
                      #shown as the “hi” and “lo” value envelopes
                      nrank = 2,
                      
                      #Suppress simulation message
                      verbose = FALSE) 


#Gather this to plot series in ggplot
gfunction_long <- gfunction %>% 
  
  #Convert as data frame
  as.data.frame() %>% 
  
  #Pivot longer
  pivot_longer(cols = obs:hi, names_to = "model", values_to = "g_val") %>% 
  
  #Recode levels
  mutate(model = recode(model, 
                        obs = "Observed Data",
                        theo = "Theoretical Complete Spatial Randomness",
                        hi = "High 95th Percentile (Left)",
                        lo = "Low 95th Percentile (Right)"))


#Plot in ggplot
ggplot(data = gfunction_long, 
       aes(x = r, y = g_val, group = model)) +
  
  #Define line plot
  #Color by model's data and change size
  geom_line(aes(color = model), size = 1.2) +

  #Specify colors
  scale_color_manual(values = c("lightblue", "lightblue",
                                "darkorange", "cyan4")) +

  
  #Change x-axis and y-axis labels
  #Change legend title
  labs(x = expression(italic("r")),
       y = expression(atop(paste(italic("G(r)")),
                           paste("(degree lat-long)"))),
       color = "Model") +
  
  #Change theme
  theme_minimal() +
  
  #Change size of ticks
  theme(axis.text = element_text(size = 10),
        
        #Bold and change size of axis labels
        axis.title = element_text(face = "bold", size = 12),
        
        #Bold legend title
        legend.title = element_text(face = "bold"),
        
        #Change legend posiiton
        legend.position = c(0.7, 0.4))
```
**Figure 4.** Plot of nearest neighbor patterns (*G(r)*) compared to a sequence of calculated distances (*r*). *G(r)* of the observed data (orange line) is above the theoretical threshold for complete spatial randomness (cyan line). Monte Carlo simulations were used to find the theoretical values of complete spatial randomness as well as the 95% upper and lower bounds (light blue lines) of the theoretical values. Data source: OSPR (2009).

<br>

Based on the density plot (**Figure 3B**) and *G(r)* plot (**Figure 4**), spatial point pattern analysis reveals that oil spill incidents in Los Angeles County from the year 2008 are clustered along similar degrees of latitude and longitude rather than spatially random. The clusters of oil spill incidents in Los Angeles County are located in, what appears to be, Long Beach, Burbank, the San Gabriel Valley, and Central Los Angeles (**Figure 3A**), which are known to be highly populated areas.


# Summary

In conclusion, we have plotted the locations of over 3,200 marine and inland oil spill incidents reported in California by different counties using `geom_sf` and `tmap`. Los Angeles County has the highest number of oil spill incidents in the year 2008, and kernel density plot and spatial pint pattern analysis using *G*-function revealed that the incidents were spatially clustered.

Future analyses could include using *L*-function in spatial point pattern analysis, and further exploring oil spill incidents at the sub-county level in Los Angeles or in other counties that have high incidents of oil spills (e.g., San Mateo County in the Bay Area).


# References

## Data and Literature

California Department of Fish and Wildlife (CDFW) (2020). California State Geoportal. Oil Spill Incident Tracking [ds394] database. https://gis.data.ca.gov/datasets/CDFW::oil-spill-incident-tracking-ds394-1/about.

California Open Data Portal. (2016). Geographic Boundaries. CA County Boundaries. https://data.ca.gov/dataset/ca-geographic-boundaries.

Office of Spill Prevention and Response (OSPR) Incident Tracking Database. (2009). Oil Spill Incident Tracking [ds394]. https://map.dfg.ca.gov/metadata/ds0394.html.


## R Libraries

Baddeley, A., Rubak, E., & Turner, R. (2015). Spatial Point Patterns: Methodology and Applications with R. London: Chapman and Hall/CRC Press, 2015. URL https://www.routledge.com/Spatial-Point-Patterns-Methodology-and-Applications-with-R/Baddeley-Rubak-Turner/9781482210200/.

Baptiste, A. (2017). gridExtra: Miscellaneous Functions for "Grid" Graphics. R package version 2.3. https://CRAN.R-project.org/package=gridExtra.

Bivand, R. & Lewin-Koh, N. (2021). maptools: Tools for Handling Spatial Objects. R package version 1.1-2. https://CRAN.R-project.org/package=maptools.

Firke, S. (2021). janitor: Simple Tools for Examining and Cleaning Dirty Data. R package version 2.1.0. https://CRAN.R-project.org/package=janitor.

Kassambara, A. (2020). ggpubr: 'ggplot2' Based Publication Ready Plots. R package version 0.4.0. https://CRAN.R-project.org/package=ggpubr.

Müller, K. (2020). here: A Simpler Way to Find Your Files. R package version 1.0.1. https://CRAN.R-project.org/package=here.

Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. The R Journal 10 (1), 439-446, https://doi.org/10.32614/RJ-2018-009.

Pedersen, T.L. (2020). patchwork: The Composer of Plots. R package version 1.1.1. https://CRAN.R-project.org/package=patchwork.

R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Tennekes, M. (2018). “tmap: Thematic Maps in R.” _Journal of Statistical Software_, *84*(6), 1-39. doi: 10.18637/jss.v084.i06 (URL: https://doi.org/10.18637/jss.v084.i06).

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686.


